\documentclass[colcode]{../common/ocaml-cs}

\lstset{language=opam}

\makeheader{The \textbf{opam} package manager for OCaml (command-line)}{opam 2.0}{1}
% warning: this is repeated for p.2

\renewcommand{\arraystretch}{1.1}
\begin{document}


\begin{multicols}{3}

\lst!opam command --help! \hfill show the manpage for \lst!command! \hfill \null

Non-ambiguous prefixes are accepted \\
(\emph{e.g.} {\lstset{morekeywords=[2]{inst}}\lst!opam inst . --deps!} for
\lst!opam install . --deps-only!).

\section{Installation}
Download from: \hfill \url{https://opam.ocaml.org} \hfill \null

\listvav{
{\lst!opam init!} &
     {set up \opam, default repository, switch, scripts} \\
\lst!opam init --bare! &
     {create \lst[language=neutral]!\~/.opam! without a compiler switch} \\
\lst!opam init --reinit -i! &
     {reinstall opam scripts (\emph{e.g.} after upgrade)} \\
}

Run \lst[language=neutral]!eval $(opam env)! when changing switch or prompted, or accept the shell hook setup. %$

\section{Configuration}
\listvav{
\lst!opam config report! & display a summary of the set-up \\
\lst!opam command -v[v]!    & print commands being run \\
\lst!opam config set v val! & set switch variable \lst!v! \\
\lst!opam config set-global v val! & set global config variable \lst!v! \\
\lst!opam command --root root! & run opam using \lst!root! as opam root \\
\lst!opam command --switch sw! & run opam on given \lst!sw! \\
\lst!opam clean! & remove archive cache and artefacts \\
}

\section{Switches}
\listvav{
\splt{\lst!opam switch create [name] compiler!}
     {install a new prefix (``switch'') with the given \lst!compiler! and select
      it. \lst!compiler! should be one of \lst!ocaml-base-compiler[.version]!,
      \lst!ocaml-system[.version]!, \lst!ocaml-variants[.version]!, or \lst!--empty!.} \\
\lst!opam switch!                 & list installed switches \\
\lst!opam switch sw!              & select the switch \lst!sw! \\
\lst!opam switch create dir [compiler]! & install packages defined in \lst!dir! in a new local switch \\
\lst!opam switch list-available!  & list all available compilers \\
\splt{\lst!opam exec [--switch sw] -- command args!}
     {run \lst!command args! in the correct environment} \\
}

The ``current switch'' is defined by the \lst!OPAMSWITCH! environment variable,
the \lst!PWD! (for local switches), and the latest selected one.

\section{Allowed URL formats}
\listvav{
\lst!http:// https:// ftp://! &
  remote archives \\
\lst!ssh:// file://! & archives or directories \\
\lst!path!                & file paths (version control is detected) \\
\lst!user@host:path!      & ssh addresses (using {\em rsync})\\
\lst!git:// hg:// darcs://! & version control \\
\linecmd{\lst!git+ssh:// hg+https:// git+file://!}
  & version control with specific transport \\
\linecmd{\lst!git+https://foo.com/git/bar\#branch!}
  & specific tag, branch, commit, etc. \\
}

\section{Packages}
\listvav{
\splt{\lst!opam install pkgs!}
     {\lst!pkgs! are package names, \lst!pkg.version!, constraints \lst!"pkg>=version"!} \\
\lst!opam install --show! & only print a list of actions \\
\lst!opam install --dry-run! & simulate everything \\
\linecmd{\lst!opam install pkgs --best-effort!} & don't fail on impossible requests \\
\lst!opam update [--all]! & update repositories and package sources \\
\lst!opam upgrade [pkgs]! & bring installed packages to latest version\\
\lst!opam remove pkgs!    & uninstall packages \\
\lst!opam remove --auto!  & uninstall no longer needed dependencies \\
\lst!opam reinstall pkgs! & recompile and reinstall packages \\
\lst!opam source pkg [--dev]! & download package source \\
% }
% \listvav{
\linecmd{\lst!opam reinstall --list-pending!} & show pending recompilations \\
\linecmd{\lst!opam reinstall --forget-pending!} & at your own risk \\
}

\section{Exploring}
\listvav{
\lst!opam list!                    & list installed packages \\
\lst!opam list --resolve pkg! &
list a {\em sufficient} set of dependencies to install \lst!pkg! \\
\lst!opam list [--rec] --required-by pkg!  & list dependencies of \lst!pkg! \\
\lst!opam list [--rec] --depends-on pkg!   & list packages depending on \lst!pkg! \\
\lst!opam list --roots!            & exclude automatically-installed dependencies \\
\lst!opam list --external pkg!     & list external \lst!pkg! dependencies \\
\lst!opam show pkg [--field=flds]! & show package details \\
\lst!opam show pkg --raw!          & show package \opam file \\
\lst!opam var v!                   & print value of opam variable \\
{\lstset{deletekeywords=[2]{list},morekeywords=[3]{list}}\lst!opam config list [pkg]!}
                                   & list variables [of package \lst!pkg!] \\
}

\section{Package pinning}
\listvav{
% \splt
{\lst!opam install dir!} &
     {pin and install packages from the sources and definitions at \lst!dir!} \\
\lst!opam pin pkg version!    & pin \lst!pkg! to given version \\
\lst!opam pin pkg[.version] url! & pin \lst!pkg!$^1$ to \lst!url! (can be a dir) and install \\
\lst!opam pin url!            & pin using package definitions at \lst!url! \\
\lst!opam pin --dev pkg!      & pin known package to its source repo \\
\lst!opam pin [--short]!      & list pinned packages \\
{\lstset{deletekeywords=[2]{remove},morekeywords=[3]{remove}}\lst!opam pin remove pkgs|dir!}
                              & unpin packages \\
\lst!opam pin edit pkg!       & tweak package definition \\
}

\lst!pin! commands also install/remove unless \lst!-n! is specified.

$^1$If not using \lst!pkg.version!, version is defined by opam
file, directory name, or latest known version.

\section{Project development}
\subsection{Working with local pins}
\listvav{
\splt{\lst!opam install pkg|dir --deps-only!}
     {just install all the pre-requisites} \\
\splt{\lst!opam install pkg|dir --working-dir!}
     {bypass VCS, take all uncommitted changes} \\
\splt{\lst!opam install pkg|dir --inplace-build!}
     {process build and install directly in the source} \\
\splt{\lst!opam install pkg|dir --assume-built!}
     {directly run install commands from the source} \\
\splt{\lst!opam lint pkg|dir|opamfile!}
     {check the style of a package definition} \\
}

\subsection{Sharing a dev setup}
\listvav{
\splt{\lst!opam lock pkg --direct-only!}
     {generate an \lst[language=neutral]!opam.locked! file with version-strict dependencies} \\
\splt{\lst!opam lock pkg!}
     {generate an \lst[language=neutral]!opam.locked! file with a fixed dependency tree} \\
\splt{\lst!opam install dir|pkg --locked!}
     {install, reproducing the same state as described by the \lst!locked! file} \\
\splt{\lst!opam switch export|import file|-!}
     {switch state (compiler, installed packages, pins\ldots) save/restore}
}

\section{Configuring remotes}
\listvav{
\splt{\lst!opam repository [--all]!}
     {list defined repositories (current switch, or all)} \\
\splt{\lst!opam switch create --repos default,custom=url ...!}
     {create a switch with repositories \lst!default!, and newly defined \lst!custom!} \\
\splt{\lst!opam repository add name url --dont-select!}
     {define repository \lst!name! at \lst!url!} \\
\splt{\lst!opam repository add name [url]!}
     {use \lst!name! in the current switch} \\
\splt{\lst!opam repository add name [url] --set-default!}
     {use \lst!name! for newly created switches} \\
\splt{\lst!opam repository add name [url] --all-switches!}
     {use \lst!name! for all existing switches} \\
\splt{\lst!opam repository add name [url] --rank=-1!}
     {use \lst!name! with lowest priority} \\
\splt{\lst!opam repository set-url name url!}
     {change repository url} \\
\splt{\lst!opam repository set-repos foo,bar!}
     {redefine the repos selections for the current switch} \\
}

The definition for \lst!pkg.version! is taken from the highest ranking
repository.

%% \section{Upgrading opam}
%% From binary: run script at
%% {\scriptsize\url{https://raw.githubusercontent.com/ocaml/opam/master/shell/install.sh}}

%% \ppar
%% From source: \\
%% \lst[language=shell]!  opam install opam-devel! \\
%% \lst[language=shell]!  sudo cp $(opam var opam-devel:lib)/opam $(which opam)!\\


\pagebreak

\makeheader{The \textbf{opam} package manager for OCaml (packaging and tools)}{opam 2.0}{1}


\section{Package definition files}
\lstset{language=opam-file}

Full specification:
\hfill\url{http://opam.ocaml.org/doc/Manual.html#opam}\hfill\null

\listvav{
In source:&\lst!opam!, or \lst!pkg.opam!, or \lst!opam/pkg.opam! \\
In a package repository:&
 {\lstset{deletekeywords=[5]version,morekeywords=[4]version}\lst!packages/pkg/pkg.version/opam!} \\
}

\begin{lstlisting}
opam-version: "2.0"
name: "project"
version: "0.1"
synopsis: "One-line description"
description: """
Longer description
"""
maintainer: "Name <email>"
authors: ["Name <email>"]
license: "SPDX license"  # see https://spdx.org/licenses/
homepage: "https://project.org"
bug-reports: "https://gitfoo.net/project/issues"
dev-repo: "git+https://gitfoo.net/project.git"
depends: ["ocaml"
          "ocamlfind" {<= "1.8"}
          "odoc" {with-doc & >= "1.0"}]
# with a regular ./configure - make
build: [["./configure" "--prefix=%{prefix}%"]
        [make]]
install: [make "install"]
# with dune (no 'install:' needed)
depends: ["dune" {>= "1.10"}] # add to your other 'depends:'
build: ["dune" "build" "-p" name "-j" jobs]
\end{lstlisting}

\columnbreak
\section{Some optional fields}
\listvav{
\lst!tags: ["org:foo" "examples"]! & for package sorting \\
\lst!depopts: [deps]!     & optional dependencies \\
\lst!substs: ["foo"]!     & expand file \lst!"foo"! from \lst!"foo.in"! \\
\linecmd{\lst!patches: ["f.patch" \{os = "macos"\}]!}
                          & conditional patches \\
\lst!run-test: [cmds]!    & only when running with \lst!--with-test! \\
\linecmd{\lst!pin-depends: [["pkg.version" "url"]]!}
  & when pinned, pin also these \\
\lst!conflicts: [deps]!   & anti-dependencies \\
\lst!available: condition!& pre-requirements \\
\lst!build-env: [CC = "foo"]!
                          & custom build/install environment \\
\linecmd{\lst!extra-source "f" \{src: "url" checksum: "sha256=..."\}!}
                          & additional downloads \\
\linecmd{\lst!post-messages: """message""" \{condition\}!}
                          & print to the user after install \\
}

When in a repository:
\smallskip
\listvav{
\lst!url \{! & \\
\lst!\ \ src: "url"! & archive URL (or VCS, in custom repos) \\
\lst!\ \ checksum: "sha512=XXX"! & supported: md5, sha256, sha512 \\
\lst!\}! & \\
}

\columnbreak
\section{Expressions}
Variables are strings, booleans or undefined values.

\listvav{
postfix conditions & \lst![make "foo"] \{condition\}! \\
dependencies & \lst@("p1" \{>= "0.5" \& != "0.7" \& condition\} | "p2")@ \\
version ordering & \lst!"1.02"! $=$ \lst!"1.2"! $<$ \lst!"1.12"! $<$ \lst!"2.0\~"! $<$ \lst!"2.0"! \\
comparisons & \lst!var = "value"!, \lst^var != ""^, \lst!var >= "0.1"! \\
interpolation & \lst!"can be \%\{var\}\% or \%\{bool-var?foo:bar\}\%"! \\
undefined & \lst!(?undef)! is \lst!false!, \lst!(undef | true)! is \lst!true!
}

\lst[language=opam]!opam var! lists the pre-defined variables\\
\lst[alsoletter=]!_:var! is \lst[alsoletter=]!pkg:var! for the current package

Some useful variables:
\smallskip
\listvav{
\multicolumn{2}{l}{\textbf{Strings}} \\
\lst!name!, \lst!version! & current package name, version \\
& allowed \emph{e.g.} as \lst!depends: ["foo" \{= version\}]! \\
\lst!lib! & this is \lst[language=neutral]!"\%\{prefix\}\%/lib"! \\
\lst[alsoletter=]!pkg:lib! & this is \lst[language=neutral]!"\%\{prefix\}\%/pkg/lib"! \\
\linecmd{\lst!arch!, \lst!os!, \lst!os-distribution!, \lst!os-family!, \lst!os-version!}
        & system detection \\
\multicolumn{2}{l}{\textbf{Booleans}} \\
\lst[alsoletter=]!pkg:dev! & \lst!pkg! was not built from a release archive \\
\lst!with-test! & tests have been enabled (package-specific) \\
\lst!with-doc! & documentation has been enabled (package-specific) \\
\lst!build! & (only in depends) don't recompile when changed \\
\lst!post! & (only in depends) not needed at build time \\
}

\end{multicols}
\vfill
\begin{multicols}{3}
\section{External dependencies}
\listvav{
\lst!name: "conf-gtk3"! & by convention, use a "conf-" prefix \\
\linecmd{\lst!depexts: ["libgtk-3-dev"] \{os-family = "debian"\}!}
                     & define system package dependencies \\
\lst!flags: conf!
          & package without install, for polling the system \\
}

Related commands:
\smallskip
\listvav{
\lst[language=opam]!opam list -e --resolve pkg!
   & print requirements of \lst[language=opam]!pkg! on this system \\
\lst[language=opam]!opam depext pkg!
  & handles requirements of \lst[language=opam]!pkg! (plugin) \\
}

\section{Publishing}
Through Github pull-requests to the official repository at \\
\dent\url{https://github.com/ocaml/opam-repository}

Automatically, using the \lst!opam-publish! plugin:
\smallskip
\listvav{
\lst[language=opam]!opam publish url!
    & publish from hosted source archive (plugin) \\
\lst[language=opam]!opam publish [dir]! & publish latest tag from detected Github origin \\
}

\section{Repository administration}
\lstset{language=opam}
To be run from the root of an opam repository:

{\lstset{keywords=[2]{admin},morekeywords=[3]{list,cache,lint}}
\listvav{
\lst!opam admin list!  & list packages \\
\lst!opam admin cache! & download all archives to cache \\
\lst!opam admin index! & generate an index (needed for HTTP) \\
\lst!opam admin lint!  & lint all packages \\
\lst!opam admin filter patterns! & only keep matching packages \\
\splt{\lst!opam admin add-constraint "pkg<=3"!}
     {add a version constraint to all dependencies towards \lst!pkg!}
}
}
%\pagebreak

\end{multicols}
\vfill
\end{document}
